Gusto: a toolkit for compatible finite element dynamical cores

### 2025-08-05 source/paper.tex
add figures

\subsection{Compressible Euler equations}
We have also implemented many of the standard test cases for the non-hydrostatic compressible Euler equations that are used in modern NWP models.
In this section we show results from a small selection of these, which demonstrate different domains in which these methods work.
Throughout this section we use the Semi-Implicit Quasi-Newton timestepper.
 \\
\\
The first test is the falling density current of \cite{straka1993numerical} in a vertical slice. This test involves the addition of diffusion to both the potential temperature and velocity equations.
Figure \ref{fig:straka} shows the final potential temperature equation for simulations using the lowest-order and next-to-lowest order finite elements on the vertical slice.
The lowest-order methods require the transport and diffusion terms to use the ``recovered'' scheme of \citet{bendall2019recovered} to attain second-order accuracy. 
\\
\begin{figure}[htp!]
\centering
\includegraphics[width=0.8\textwidth]{straka.png}
\caption{The potential temperature field after 900 s from the falling density current test of \cite{straka1993numerical}. (Top) with the lowest-order finite elements, and (bottom) with the next-to-lowest-order finite elements.
}
\label{fig:straka}
\end{figure}
\\
To demonstrate the use of moist physics parametrisations, we use a moist three-dimensional rising thermal simulation in an extruded doubly-periodic plane.
The initial conditions and moisture parametrisation are those of \citep{bendall2020compatible}, which is an extension of the dry case of \citep{kelly2012continuous} and \citep{melvin2019mixed}.
Cross-sections of the initial and final equivalent potential temperature field are shown in Figure \ref{fig:3d_bubble}.
As in the two-dimensional test of \citep{bryan2002benchmark}, the atmosphere is saturated and cloudy everywhere, so that the moist processes are active in the whole domain.
\\
\begin{figure}[htp!]
\centering
\includegraphics[width=0.9\textwidth]{moist_bryan_fritsch.png}
\caption{Vertical cross-sections of the (left) initial and (right) final equivalent potential temperature field from the three-dimensional moist rising thermal test.
}
\label{fig:3d_bubble}
\end{figure}
\\
Our final test case is the deep baroclinic wave of \citep{ullrich2014proposed} on the sphere.
An initial state in thermal wind balance is perturbed to produce a mid-latitude baroclinic instability, providing a flow that is typical of those found in Numerical Weather Prediction.
The surface pressure and temperature fields after 10 days are shown in Figure \ref{fig:baroclinic}, which compare well with those in \citep{melvin2024mixed} and demonstrate that our methods can replicate results from the literature.

>>>>

\subsection{Compressible Euler equations}
We have also implemented many of the standard test cases for the non-hydrostatic compressible Euler equations that are used in modern NWP models.
In this section we show results from a small selection of these, which demonstrate different domains in which these methods work.
Throughout this section we use the Semi-Implicit Quasi-Newton timestepper.
 \\
\\
The first test is the falling density current of \cite{straka1993numerical} in a vertical slice. This test involves the addition of diffusion to both the potential temperature and velocity equations.
Figure \ref{fig:straka} shows the final potential temperature equation for simulations using the lowest-order and next-to-lowest order finite elements on the vertical slice.
The lowest-order methods require the transport and diffusion terms to use the ``recovered'' scheme of \citet{bendall2019recovered} to attain second-order accuracy. 
\\
\begin{figure}[htp!]
\centering
\includegraphics[width=0.8\textwidth]{straka.png}
\caption{The potential temperature field after 900 s from the falling density current test of \cite{straka1993numerical}. (Top) with the lowest-order finite elements, and (bottom) with the next-to-lowest-order finite elements.
}
\label{fig:straka}
\end{figure}
\\
To demonstrate the use of moist physics parametrisations, we use a moist three-dimensional rising thermal simulation in an extruded doubly-periodic plane.
The initial conditions and moisture parametrisation are those of \citep{bendall2020compatible}, which is an extension of the dry case of \citep{kelly2012continuous} and \citep{melvin2019mixed}.
A cross-sections of the final equivalent potential temperature field is shown in Figure \ref{fig:3d_bubble}.
As in the two-dimensional test of \citep{bryan2002benchmark}, the atmosphere is saturated and cloudy everywhere, so that the moist processes are active in the whole domain.
\\
\begin{figure}[htp!]
\centering
\includegraphics[width=0.48\textwidth]{moist_3d_bubble.png}
\caption{Vertical cross-section of the final equivalent potential temperature field from the three-dimensional moist rising thermal test.
}
\label{fig:3d_bubble}
\end{figure}
\\
Our final test case is the deep baroclinic wave of \citep{ullrich2014proposed} on the sphere.
An initial state in thermal wind balance is perturbed to produce a mid-latitude baroclinic instability, providing a flow that is typical of those found in Numerical Weather Prediction.
The surface pressure and temperature fields after 10 days are shown in Figure \ref{fig:baroclinic}, which compare well with those in \citep{melvin2024mixed} and demonstrate that our methods can replicate results from the literature.
